Higher-Order unification for free!

### 2024-04-15 paper/main.tex
sec 5.2 and 5.3 until lemma 5.11 + example eta-progress-lhs

\noindent
TWhenever \elpiIn{flam F} is detected to be in
\maybeeta it is compiled to \elpiIn{lam F1} and replaced by the fresh
variable \elpiIn{A}. This variable sees all the names free in
\elpiIn{lam F1}. The variable is connected to \elpiIn{lam F1} via a \linketa.
Invariant~\fullref{inv:linklhs} holds for this link, moreover:

@@

\begin{proof}[Proof sketch]
  By contradiction, suppose that the rule above is applied and that
  the \rhs of the link is $\lambda x.\lambda y.t$, where $x$ and $y$ occur in $t$.
  If $\testmaybeeta{}~\lambda y.t$ holds then the recursive call to
  \elpiIn{comp} (made by \elpiIn{comp-lam}) must have put a fresh variable
  in its place, so this case is impossible.
  Otherwise, if $\testmaybeeta{}~\lambda y.t$ does not hold, also
  $\testmaybeeta{}~\lambda x.\lambda y.t$ does not hold, contradicting
  the assumption that the rule was applied.
\end{proof}

@@

\newcommand{\progressetaright}{\emph{$\eta$-progress-\rhs}\xspace}
\begin{definition}[\progressetaright]\label{def:progressetaright} A link
\linketaM{\Gamma}{X}{T} is removed from \linkStore when either 1)
$\testmaybeeta~T$ does not hold (anymore) or 2) by $\eta$-contracting $T$ to
$T'$ where $T'$ does not start with the \elpiIn{lam} constructor. In the
first case, $X$ is unified with $T$ and in the second one, $X$ is unified with
$T'$ (under the context $\Gamma$).
\end{definition}

@@

\newcommand{\progressetadedup}{\emph{$\eta$-progress-deduplicate}\xspace}
\begin{definition}[\progressetadedup]\label{def:progressetadedup}
  A link \linketaM{\Gamma}{X_{\vec{s}}}{T} is removed from \linkStore when
  another link \linketaM{\Delta}{X_{\vec{r}}}{T'} is in  \linkStore.
  By \cref{inv:uvaarity} the length of $\vec{s}$ and $\vec{r}$ is the same
  hence we can move the term $T'$ from $\Delta$ to $\Gamma$ by renaming its
  bound variables, i.e. $T'' = T'[\vec{r}/\vec{s}]$.
  We then run $T \Ue{} T''$ (under the context $\Gamma$).
\end{definition}

@@

The example at the beginning of \cref{sec:eta}, once
$\sigma = \{~ A \mapsto f ~\}$, triggers \progressetaleft since the link
becomes \linketaM{}{f}{\lambda x.B_{x}} and the \lhs is a constant.
In turn the rule \todo{which rule} runs $\lambda x.f \appsep x \Ue{} \lambda x.B_{x}$,
resulting in $\sigma = \{~ A \mapsto f ~;~ B_{x} \mapsto f ~\}$.
Decompilation the generates $\rho = \{~ X \mapsto f ~\}$, since
$X$ is mapped to $B$ and
$f$ is the $\eta$-contracted version of $\lambda x.f \appsep x$.
\todo{explain why decompile does so and which eta?}

>>>>

\noindent
Whenever \elpiIn{flam F} is detected to be in
\maybeeta it is compiled to \elpiIn{lam F1} and replaced by the fresh
variable \elpiIn{A}. This variable sees all the names free in
\elpiIn{lam F1}. The variable is connected to \elpiIn{lam F1} via a \linketa.
Invariant~\fullref{inv:linklhs} holds for this link, moreover:

@@

\begin{proof}[Proof sketch]
  By contradiction, suppose that the rule above is applied and that
  the \rhs of the link is $\lambda x.\lambda y.t$, where $x$ and $y$ occur in $t$.
  If $\testmaybeeta{}~\lambda y.t$ holds then the recursive call to
  \elpiIn{comp} (made by \elpiIn{comp-lam}) must have put a fresh variable
  in its place, so this case is impossible.
  Otherwise, if $\testmaybeeta{}~\lambda y.t$ does not hold, then also
  $\testmaybeeta{}~\lambda x.\lambda y.t$ does not hold either, contradicting
  the assumption that the rule was applied.
\end{proof}

@@

\newcommand{\progressetaright}{\emph{$\eta$-progress-\rhs}\xspace}
\begin{definition}[\progressetaright]\label{def:progressetaright} A link
\linketaM{\Gamma}{X}{T} is removed from \linkStore when either 1)
$\testmaybeeta~T$ does not hold (anymore) or 2) by $\eta$-contracting $T$ to
$T'$ where $T'$ does not start with the \elpiIn{lam} constructor. In the
first case, $X$ is unified with $T$, and in the second one, $X$ is unified with
$T'$ (under the context $\Gamma$).
\end{definition}

@@

\newcommand{\progressetadedup}{\emph{$\eta$-progress-deduplicate}\xspace}
\begin{definition}[\progressetadedup]\label{def:progressetadedup}
  A link \linketaM{\Gamma}{X_{\vec{s}}}{T} is removed from \linkStore when
  another link \linketaM{\Delta}{X_{\vec{r}}}{T'} is in  \linkStore.
  By \cref{inv:uvaarity} the length of $\vec{s}$ and $\vec{r}$ is the same;
  hence we can move the term $T'$ from $\Delta$ to $\Gamma$ by renaming its
  bound variables, i.e. $T'' = T'[\vec{r}/\vec{s}]$.
  We then run $T \Ue{} T''$ (under the context $\Gamma$).
\end{definition}

@@

The example at the beginning of \cref{sec:eta}, once
$\sigma = \{~ A \mapsto f ~\}$, triggers \progressetaleft since the link
becomes \linketaM{}{f}{\lambda x.B_{x}} and the \lhs is a constant.
This rule runs $\lambda x.f \appsep x \Ue{} \lambda x.B_{x}$,
resulting in $\sigma = \{~ A \mapsto f ~;~ B_{x} \mapsto f ~\}$.
By the definition of \linketa decompilation,
decompilation generates $\rho = \{~ X \mapsto \lambda x. f \appsep x ~\}$, since:
$X$ is mapped to $B_x$ and $B_x$ is decompiled into $\lambda x.f \appsep x$.
% \todo{explain why decompile does so and which eta?}
